# Data 
```{r setup, warning=FALSE}
library(readr)
library(dbplyr)
library(tidyverse)
```

## Sources
We collected all of our data from the [chess.com game database](https://www.chess.com/games).
Where exactly all of this site acquired its data is a difficult paper trail to
follow.

Organized chess games have been recorded for a long time, long before
the internet or the Turing machine had even been concieved. One common instance
is that originally, players in an event would write down the game in terms of
moves on a slip of paper, provide it to the tournament organizer, and then that
organizer would store that in some other public record. How exactly chess.com
has procured it all doesn't seem to be specified--and it is also the source
of most of our inconsistent or "missing" data. Some of it even predates (or
in was ignorance of) the ELO system, one of our selected features.

As for the modern day, many of these games are in fact played on this very
platform, giving a much greater coherency, consistency, and volumn to the data.
So when modern-day players have far more games and a more consistent track
record, that is why.

[Chess.com]((https://www.chess.com) generously provides download buttons for every
page of games found (up to 25 games per page). While there is a download button
for the whole page, there is unfortunately not a download for all pages at once.
So to acquire our whole dataset, we had to scrape each players page (which
arguably is allowed under their [robots.txt](https://www.chess.com/robots.txt),
as we are not using the `games/search` pages, though it does redirect to them...),
and write a parser to conjoin all of the data per player. Both the scraper and
parser were written in Python.

As chess is an everlasting game, and as there have been an unknown number of
players and organizers, there are many, many different features that come and
go from individual records. The primary commonalities are the variables:
`Event`, `White` and `Black` (player colors), `WhiteElo` and `BlackElo`, and
`Result` (`1-0`, `0-1`, or `1/2-1/2` for draws), two IDs for chess.com, and of 
course the list of moves. All were stored in the format `[<name> "<value>"]`,
except for the most important feature(s), the moves. In the `.pgn` files the 
moves are a separate type offield that is essentially a list; we needed to parse
and break that into individual moves per move number. As a game can be any 
number of moves, we broke that up into one move (and one move number) per row,
and selected all of the other features as mentioned, aside from `Event` and the
ids. For the purposes of our analysis, no other feature was necessary, though an
analysis of time could be interesting. Every `.pgn` file had 25 games until the
final page per player, and the actual number of pages varied highly by player.
In total, there were 39309 games in our dataset, ranging from ~500 to ~2400 
games per player. Additionally, since we're analyzing the opening moves, and 
white always plays first, per player in our analysis we only scraped the games 
where they played as white.

Naturally, in a dataset this varied, there were a few problems to be handled by
our parser. One, clocktime per move was occasionally included in the games' move
area, warping every move from a format like `1. d4 d5 2. c4 e6 ...` to 
`1. d4 {[%clk 0:15:20]}  Nf6 {[%clk 0:15:20]} 2. c4 {[%clk 0:15:28]}  e6 {[%clk 0:15:29]} ...`,
which we had to filter out. Additionally, three games had a result of `*`
instead of a record. After tracking them down, they were all aborted games that
were treated as draws; they all were played on the internet, as well, though we
aren't certain as to the precise cause of the aborts.

## Cleaning / transformation
Our cleaning process was all done in the Python parser `pgn_parser.py`. It all
may as well happened there, since the parser needed to be equipped to read all
of the data files anyway (it'd crash otherwise). As such, all of the issues 
mentioned in 3.1 were handled in the parser. To tidy the dataset, also used the
parser and broke the single field of moves into one move and one move number per
row. As a result, there is no unique identifier per row, and in fact there isn't
one per game either, as some players played one another many times. We created 
one `.csv` per player, then a whole conjoined `.csv` for ease of loading into a 
unifying dataframe.


## Missing value analysis
Depending on how you define it, there is an inordinate amount of missing data
here--only games that have been recorded could possible be featured, and with
modern game playforms modern players will naturally have a much more complete
record. However, aside from simply summing the games per era per player, it's not
possible to analyze that. Instead, we are more interested in one other sometimes
missing field: the players ELOs.

There are three types of value here that represents missing, `0`, `?`, and `N/A`.
In parsing, we set all of those to `?`, but there isn't much more we can do to
resolve them. Some games took place before ELO was introduced, others were human
error or simply deemed unimportant, and since ELO is a constantly shifting
calculation, there's no way to perfectly recreate the number. However, the
players do always have an ELO, and between the three types of missing,
it seems to be by convention and not by any deeper meaning. Therefore, we simply
count the number of missing by player.


```{r}
# Get a list of files that match a certain pattern
files <- list.files(path="player-moves", pattern="*csv", full.names=TRUE, recursive=FALSE)

all_moves <- read.csv("player-moves/all-games.csv")
colnames(all_moves) <- c('game_index', 'white_name', 'white_elo', 'black_name', 'black_elo', 'result', 'move_number', 'move')

all_games = all_moves |>
  filter(move_number == 1)

games_by_player = all_games |>
  group_by(white_name) |>
  summarize(total_count = n())

games_missing_elo = all_games |>
  filter(white_elo == '?' | black_elo == '?')

missing_by_player = games_missing_elo |>
  group_by(white_name) |>
  summarize(total_count = n())

missing_by_player['frequency'] = missing_by_player['total_count'] / games_by_player['total_count']

missing_by_player |>
  mutate(white_name = fct_reorder(white_name, desc(frequency), .desc = TRUE)) |>
  ggplot(aes(x = white_name, y = frequency)) +
    geom_bar(stat="identity") +
    xlab("Player") +
    ylab("Missing ELO Frequency") +
    coord_flip()
```

As expected, the players with the most (and sometimes complete) missing ELO 
values are the earlier champions, many of whom were active before the advent of
the ELO system, while the players with the least missing are all active in the
modern, electronically recorded era.

Another point of interest: of the games that have one or more errors, within
the records, what is the distribution of white-to-black missing ELOs?

```{r}

games_missing_elo$elo_missing <- as.factor(ifelse(games_missing_elo$white_elo != "?", "Black",
                                           ifelse(games_missing_elo$black_elo != "?", "White", "Both")))
head(games_missing_elo)

ggplot(data = games_missing_elo, mapping = aes(elo_missing)) +
  xlab("Player(s) Missing ELO") +
  ylab("Missing Count") +
  geom_bar()
```

So even though--since they are world champions, or less commonly grandmasters--
the white player is somewhat more commonly the one missing ELO. Of course, it is
unsurprisingly most common that both miss their ELO, again indicating records
that didn't notate ELO, or simply predated its use.